Choosing Color Palettes for Statistical Graphics

### 2007-01-12 trunk/vcd/paper/Color/TAS.Rnw
first TAS revision finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@638 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose a suitable palette for a certain visualization task. Consequently,
sub-optimal color palettes are used for too many color graphics displayed in journal articles
or on presentation slides which could easily be enhanced if the underlying information
were captured appropriately in the color schemes. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

@@

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
extremely easy to implement in software and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006} and the HCL
color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2006}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
several motivating examples, showing how typical HSV-based graphics can be enhanced
by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
Section~\ref{sec:software} offers some general remarks on the implementation in statistical
software packages as well as some details on our implementation in \proglang{R}.
Section~\ref{sec:discussion} concludes the paper with a discussion.

@@

The simplest solution to this task is to employ light/dark contrasts, i.e., rely
on the most basic perceptual axis. The interestingness can be coded
by an increasing amount of gray corresponding to decreasing luminance in HCL space:
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $i' = f(i)$ that
controls whether luminance is increased quickly with intensity or not. We found $i' = f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

@@

The \proglang{R} system for statistical computing \citep{vcd:R:2006} already comes
with an open-source implementation of HCL (and other color spaces) in the package
\pkg{colorspace} \citep{vcd:Ihaka:2006}. The coordinate transformations mentioned
above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
port to other statistical software systems. Based on \pkg{colorspace}, 
the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} provides implementations
of all palettes discussed above. Qualitative palettes are provided by
\code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
\proglang{R}). Sequential palettes based on a single hue are implemented in the function
\code{sequential_hcl()} while \code{heat_hcl()} offers sequential palettes based on
a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
Technical documentation along with a large collection of examples is available via
\verb/help("rainbow_hcl")/. \proglang{R} code for reproducing
Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "vcd")}.

@@

We provide paths through HCL space along perceptual axes
so that colors selected along these paths match perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:motivation}), chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance.
%% Another parameter in these 
%% palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
%% Employing a function $|i|^p$ for this, small
%% powers $p$ can be used if palettes with a lot of color should be 
%% constructed (e.g., to clearly separate regions in a classification map). On
%% the other hand, large powers $p$ will result in palettes with less color
%% (e.g., appropriate to highlight only extreme regions firmly classified in such a map).

>>>>

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose a suitable palette for a certain visualization task. Consequently,
sub-optimal color palettes are used quite frequently for color graphics displayed in journal articles
or on presentation slides which could easily be enhanced if the underlying information
were captured appropriately in the color schemes. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

@@

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space.
%% Given a mapping from HCL to RGB coordinates, our formulas are
%% extremely easy to implement in software and we provide such an implementation
%% in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
%% system \citep[see][]{vcd:Murrell:2006} and the HCL
%% color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2006}. This gives the user
%% both a conceptual and computational tool box for experimenting with color palettes for
%% a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
several motivating examples, showing how typical HSV-based graphics can be enhanced
by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
Section~\ref{sec:software} offers some general remarks on the implementation in statistical
software packages as well as some details on the implementation in \proglang{R}.
Section~\ref{sec:discussion} concludes the paper with a discussion.

@@

The simplest solution to this task is to employ light/dark contrasts, i.e., rely
on the most basic perceptual axis. The interestingness can be coded
by an increasing amount of gray corresponding to decreasing luminance in HCL space:
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $i' = f(i)$ that
controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

@@

The \proglang{R} system for statistical computing \citep{vcd:R:2006} already comes
with an open-source implementation of HCL (and other color spaces) in the package
\pkg{colorspace} \citep{vcd:Ihaka:2006}. The coordinate transformations mentioned
above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
port to other statistical software systems. Based on \pkg{colorspace} and
the \proglang{R} graphics system \citep[see][]{vcd:Murrell:2006},
the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} provides implementations
of all palettes discussed above. Qualitative palettes are provided by
\code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
\proglang{R}). Sequential palettes based on a single hue are implemented in the function
\code{sequential_hcl()} while \code{heat_hcl()} offers sequential palettes based on
a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
Technical documentation along with a large collection of examples is available via
\verb/help("rainbow_hcl", package = "vcd")/. \proglang{R} code for reproducing
Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "vcd")}.

@@

We provide paths through HCL space along perceptual axes of the human visual system
so that colors selected along these paths match human perceptual dimensions.
This gives the users the possibility to conveniently experiment with the HCL-based
palettes by varying several graphical parameters. For qualitative
palettes, these are the coordinates on the chroma and luminance axis, respectively,
controlling whether the colors are light or dark and how colorful they are.
For sequential and diverging palettes, the user can decide whether contrasts
in the chroma or luminance direction (or both) should be employed. In our
experience (as illustrated in Section~\ref{sec:motivation}), chroma contrasts
work sufficiently well if a small set of colors is used. However, when a larger
set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
it is much more important to have a big difference in luminance.
%% Another parameter in these 
%% palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
%% Employing a function $|i|^p$ for this, small
%% powers $p$ can be used if palettes with a lot of color should be 
%% constructed (e.g., to clearly separate regions in a classification map). On
%% the other hand, large powers $p$ will result in palettes with less color
%% (e.g., appropriate to highlight only extreme regions firmly classified in such a map).
